Left-linear rules in adhesive categories

### 2024-02-10 icalp/fusione.tex
minor things

When working with left-linear rewriting systems, sequential independence does not imply 
switchability (while the converse implication clearly holds), as shown by the example below.

\begin{example}
  \label{ex:diff1}
  Recall that, given two posets $(P, \leq)$ and $(Q, \leq)$, we can
  construct the poset $(P, \leq)\oplus(Q,\leq)$ which has as
  underlying set the disjoint union $P+Q$ and in which every element
  of $Q$ is greater than every element of $P$, while elements of $P$
  and $Q$ are compared using the original orders.

  Take now the set $S:=\{a,b,c\}$ and endow it with the smallest order
  such that $a\leq b$ and $a\leq c$.  Take also the set $\mathbb{N}$
  ordered by $\geq$, so that $0$ is its maximum.  Let $\X$ be the
  category associated to the order
  $(S, \leq)\oplus (\mathbb{N}, \geq)$, which by \Cref{rem:iso} is
  $\mathsf{Iso}$-adhesive. Notice that in this category the arrows
  $(a,b)\colon a\to b$ $(a,c)\colon a\to c$ do not have a pushout
  because $b$ and $c$ do not have a supremum.  Define $\R$ as the set
  containing the following two rules
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the following derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ of length $2$, made by two
  sequential independent direct derivations
  \[
    \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
      \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
      \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
      \ar@{>->}[ll]^{(\id{c})} c \ar[rr]_{(c,0)}&& 0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]

  \noindent
  \parbox{10cm}{ \hspace{15pt}On the other hand, the rule
    applied by $\dder{D}_1$ cannot be applied to $c$. indeed the
    unique morphism $a\to c$ is $(a,c)$, thus we get a diagram
    aside. But $b$ and $c$ do not have a supremum in the poset
    underlying $\X$, thus we do not get a direct derivation from
    $c$.}  \parbox{3cm}{
    \vspace{-.3cm}$\xymatrix{a \ar[d]_{(a,c)}& a
      \ar[d]_{(a,c)}\ar@{>->}[l]_{\id{a}} \ar[r]^{(a,b)} & b \\c
      & c \ar[l]^{\id{c}} }$}  
\end{example}

@@

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback.
  %given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the resulting squares, depicted  
  below, are pushouts and the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix@R=5pt{
      %
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
      % &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
      % & P \ar@{.>}[r]_{q_1} &
      Q_1}
  \]
\end{definition}

The category $\cat{Graph}$ is tame by the result in ..., while
example~\ref{ex:diff1} shows an adhesive non-tame category.

@@

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Tame rewriting systems]A left-linear DPO rewriting system is \emph{tame} if every independence pair between two direct derivations is good.
\end{definition}
\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not good, and thus of a non tame
  rewriting system.
\end{example}

>>>>

@@

Unfortunately, when working with left-linear rewriting systems,
sequential independence does not imply switchability (while the
converse implication clearly holds), as shown by the example below.

\begin{example}
  \label{ex:diff1}
  Given two posets $(P, \leq)$ and $(Q, \leq)$, define
  $(P, \leq)\oplus(Q,\leq)$ as the poset which has as underlying set
  the disjoint union $P+Q$ and in which every element of $Q$ is
  greater than every element of $P$, while elements of $P$ and $Q$ are
  compared using the original orders.

  Take now the set $S:=\{a,b,c\}$ and endow it with the smallest order
  such that $a\leq b$ and $a\leq c$.  Take also the set $\mathbb{N}$
  ordered by $\geq$, so that $0$ is its maximum.  Let $\X$ be the
  category associated to the order
  $(S, \leq)\oplus (\mathbb{N}, \geq)$, which by \Cref{rem:iso} is
  $\mathsf{Iso}$-adhesive. Notice that in this category the arrows
  $(a,b)\colon a\to b$ $(a,c)\colon a\to c$ do not have a pushout
  because $b$ and $c$ do not have a supremum.  Consider a rewriting
  system with set of rules $\R$ containing the following two rules
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ of length $2$. 
  \[
    \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
      \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
      \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
      \ar@{>->}[ll]^{(\id{c})} c \ar[rr]_{(c,0)}&& 0 &&
      \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]

  \noindent
  \parbox{10cm}{ \hspace{15pt} Notice that the two
    direct derivations are sequential independent.
    %
    On the other hand, the rule
    applied by $\dder{D}_1$ cannot be applied to $c$. In fact, the
    unique morphism $a\to c$ is $(a,c)$, thus we get a diagram
    aside. But $b$ and $c$ do not have a supremum in the poset
    underlying $\X$, thus we do not get a direct derivation from
    $c$.}  \parbox{3cm}{
    \vspace{-.3cm}$\xymatrix{a \ar[d]_{(a,c)}& a
      \ar[d]_{(a,c)}\ar@{>->}[l]_{\id{a}} \ar[r]^{(a,b)} & b \\c
      & c \ar[l]^{\id{c}} }$}  
\end{example}

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback.
  %given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the resulting squares, depicted  
  below, are pushouts and the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix@R=5pt{
      %
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
      % &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
      % & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Tame rewriting systems]A left-linear DPO rewriting system is \emph{tame} if every independence pair between two direct derivations is good.
\end{definition}

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not good, and thus of a non tame
  adhesive rewriting system. Instead the category $\cat{Graph}$ is
  tame by the result in ....
\end{example}
